Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-11-12 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Update draft (restructure and fixmes)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2865 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 39f5f0cd12ef4f88b3e71d82673e607ddb4ac692

\numberwithin{equation}{subsection}

@@

\title{Distributional Trees and Forests for Circular Data with an Application to Probabilistic Wind Direction Forecasting}
\Shorttitle{Distributional Trees and Forests for Circular Data}

@@

\Abstract{
As circular data can be found in many different subject areas accounting for periodicity has 
become an important aspect in establishing appropriate statistical models.
For probabilistic modeling of circular data the von Mises distribution is widely used.
To capture how its parameters change with covariates, a regression tree model
is proposed as an alternative to more commonly-used additive models. The resulting
distributional trees are easy to interpret, can detect non-additive effects, and select
covariates and their interactions automatically. In order to stabelize and regularize
the model an ensemble of distributional trees is combined yielding a distributional
forest which allows for modeling smooth effects as well.
For illustration, hourly wind direction forecasts are obtained at two airports in Austria
(Vienna and Innsbruck) based on a set of meteorological measurements from each of the 
airports and surrounding weather stations.
}
\Keywords{Distributional Trees and Forests; Circular Response; Von Mises Distribution}

@@

@@

Therefore, we propose a very flexible alternative approach by adapting regression trees and random 
forests to circular data, specifically by employing the von Mises distribution within the methodology
of distributional trees and distributional forests following \cite{Schlosser+Hothorn+Stauffer:2019}. 
In that way no model assumptions are required as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while combining an 
ensemble of trees to a forest model allows for modeling smooth changes as well. 
Thus, a high level of flexibility is provided while a full distributional model is specified 
by distributional trees and forests such that a wide range of inference is made available.
The corresponding methodology and its features are presented in Section~\ref{sec:tree_forest} after
discussing distributional models for circular data in Section~\ref{sec:prob_circ}.
In Section~\ref{sec:wind} the presented models are applied to probabilistic forecasting of wind direction
at the airports of Innsbruck and Vienna.

\section{Probabilistic Circular Modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper circular
probability distribution, i.e., a probability distribution which accounts for the periodicity
of circular data.
Generally, this feature can be obtained by ``wrapping'' the probability density function
of any continuous distribtuion around the unit circle \citep{Mardia+Jupp:2009}, for example by
applying the modulo operation on the interval length $2\pi$.
In that way, the wrapped Cauchy distribution or the wrapped normal distribution can be employed to model symmetric unimodal circular data.
A close approximation to the wrapped normal distribution that is mathematically simpler and hence easier 
to use \citep{Fisher:1993} is provided by the von Mises distribution,
a purely circular distribution which is also known as ``the circular normal distribution''
and commonly applied for probabilistic modeling of circular data.
Based on a location parameter $\mu \in (-\pi, \pi]$ and a concentration parameter $\kappa > 0$ 
the density of the von Mises distribution for an observation $y \in (-\pi, \pi]$ is given by:
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$
\citep[see, e.g.,][for a more detailed overview]{Jammalamadaka+Sengupta:2001}.

@@

\begin{figure}[t]
\centering
\includegraphics[width=4cm]{density_vM.pdf}
\caption{Illustration of a distributional model fitted to circular data by maximum likelihood-estimation
employing the von Mises distribution. Empirical histogram (gray) and fitted density (red line) are depicted
along with the estimated location parameter (red hand).
}
\label{fig:density_vM} 
\end{figure}

@@

\section{Regression Trees and Forests for Circular Data}
\label{sec:tree_forest}
%\subsection{Motivation}
A common way of including covariates as regressors in distributional models is to apply
generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} where each distribution parameter is estimated
by its own generalized additive model (GAM).
As already discussed in Section~\ref{sec:introduction}, additive strategies work well 
for smooth effects but difficulties can arise in case of non-additive changes. Additionally, 
specifying a model properly can be very challenging, particularly for a high number of covariates 
and no information on possible interactions.
An alternative regression approach that can deal with these problems is to apply a tree-structured 
model.

\subsection{Distributional Trees for Circular Data}
\label{sec:circtree}
Fitting a global model to a full data set can be very challenging, particularly for 
complex data with strong variations. Therefore, separating the data set into more homogeneous 
subgroups based on covariates before fitting a local model in each of these subgroups allows 
to capture (possibly) group specific effects more precisely and hence can result in a better 
fitting overall model.
This is the general idea of regression trees which are combined with distributional regression
modeling in \cite{Schlosser+Hothorn+Stauffer:2019} based on the unbiased recursive partitioning 
algorithms MOB \citep{Zeileis+Hothorn+Hornik:2008} or CTree \citep{Hothorn+Hornik+Zeileis:2006}.
By applying the presented distributional trees a full distributional model is specified in each 
node of the tree. For this purpose a distribution family has to be chosen in advance. Hence, 
selecting the von Mises distribution allows for an application of distributional trees to circular
data following the tree building algorithm as introduced in Section~2.2. of 
\cite{Schlosser+Hothorn+Stauffer:2019}.

@@

\subsection{Distributional Forests for Circular Data}
\label{sec:circforest}
A natural extension of distributional trees for circular data are ensembles or forests 
that can improve forecasts by regularizing and stabilizing the model.
Random forests introduced by \cite{Breiman:2001} average the predictions of an ensemble
of trees, each built on a subsample or bootstrap of the original data. 
A generalization of this strategy is to obtain weighted predictions by adaptive local 
likelihood estimation of the distributional parameters 
\citep[Section~2.3. of][]{Schlosser+Hothorn+Stauffer:2019, Hothorn+Zeileis:2017}. 
More specifically, for each possibly new observation~$\bm{z}$ a set of averaged ``nearest neighbor''
weights $w_i^{forest}(\bm{z})$ 
%\citep{Lin+Jeon:2006}
is obtained that is based on the number of trees in which $\bm{z}$ is assigned to the same terminal 
node as each learning observation $y_i, i \in \{1,\ldots,n\}$.
Hence, for a forest of $T$ trees the weights are calculated via
\begin{equation}
w^{\text{forest}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
\frac{\mathbf{1}((\bm{z}_i \in \mathcal{B}^t_b) \land (\bm{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
\end{equation}
where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
segment of the $t$-th tree.
Hence, similar observations ending up more often in the same terminal node have higher weights
and in that way a stronger influence in the estimation process.

@@

\section{Case Study: Wind Direction Forecasts}
\label{sec:wind}

@@

\subsection{Motivational Example}\label{sec:mot_ex}
As a first illustration Figure~\ref{fig:tree_ibk} displays a distributional
tree applied to predict the wind direction at the airport of Innsbruck in the following hour. 
Innsbruck lies at the bottom of a deep valley in the Alps. Topography channels wind along the 
west-east valley axis or along a tributary valley intersecting from the south. Hence, pressure 
gradients to which valley wind regimes react both west and east of the airport are considered 
as covariates along with other meteorological measurements at the airport (lagged by one hour),
such as wind direction and wind speed at Innsbruck Airport.

\begin{figure}[t]
\centering
%\includegraphics[height = .5\textheight,angle=90,origin=c]{schlosserfig1.pdf}
\includegraphics[width = \textwidth]{tree_ibk.pdf}
\caption{Fitted tree based on the von Mises distribution for wind direction forecasting.
In each terminal node the empirical histogram (gray) and fitted density (red line)
are depicted along with the estimated location parameter (red hand). The covariates
employed are wind direction (degree), wind speed ($\text{ms}^{-1}$),
and pressure gradients ($\text{dpressure; hPa}$) west and east of the airport,
all lagged by one hour.}
\label{fig:tree_ibk} 
\end{figure}

@@

\fixme{information on NAs}

@@

\subsection{Evaluation Setting and Models Applied}
Along with distributional trees and forests for circular data three other models have been applied
in this study. In particular, they have been evaluated in a five-fold cross validation with each of 
the five years contained in the data set being used as testing data once after the models where 
learned on the remaining four years.

@@

@@

\fixme{Comment on aggregation}

@@

@@

\fixme{references}

@@

\fixme{information on applied implementation here or in separate chapter?}

\subsection{Results}
In order to test and compare the models over varying settings not only 1-hour but also 3-hour
forecasts are considered at both airports yielding four different scenarios of which the results
are discussed in this section.

@@

\begin{figure}[t]
\centering
\setkeys{Gin}{width=1.0\textwidth}
%\hspace{-2cm}
\includegraphics{boxplot_crpsraw.pdf}
\caption{}
\label{fig:boxplot_crpsraw} 
\end{figure}

@@

\begin{figure}[t]
\centering
\setkeys{Gin}{width=1.0\textwidth}
%\hspace{-2cm}
\includegraphics{boxplot_crpsskill.pdf}
\caption{}
\label{fig:boxplot_crpsskill} 
\end{figure}

\section{Discussion and Outlook}
Distributional trees for circular responses are established by coupling
model-based recursive partitioning with the von Mises distribution.
The resulting trees can capture nonlinear changes, shifts, and potential interactions
in covariates without prespecification of such effects. This is particularly
useful for modeling wind direction in mountainous terrain where wind shifts
can occur due to turns of the pressure gradients along a valley.

\fixme{Summary of results}

\fixme{Possible extensions of models, use of NWP data}

>>>>

\numberwithin{equation}{section}

@@

\title{Circular Regression Trees and Forests with an Application to 
    Probabilistic Wind Direction Forecasting}
\Shorttitle{Distributional Trees and Forests for Circular Data}

@@

\Abstract{
As circular data can be found in many different subject areas accounting for periodicity has 
become an important aspect in establishing appropriate statistical models.
For probabilistic modeling of circular data the von Mises distribution is widely used.
To capture how its parameters change with covariates, a regression tree model
is proposed as an alternative to more commonly-used additive models. The resulting
distributional trees are easy to interpret, can detect non-additive effects, and select
covariates and their interactions automatically. In order to stabilize and regularize
the model an ensemble of distributional trees is combined yielding a distributional
forest which allows for modeling smooth effects as well.
For illustration, hourly wind direction forecasts are obtained at two airports in Austria
(Vienna and Innsbruck) based on a set of meteorological measurements from each of the 
airports and surrounding weather stations.
}
\Keywords{Distributional Trees and Forests; Circular Response; Von Mises Distribution}

@@

\fixme{revise literature survey with focus on transition to GLM}

@@

\fixme{revise transition to trees/forests}

Therefore, we propose a very flexible alternative approach by adapting regression trees and random 
forests to circular data, specifically by employing the von Mises distribution within the methodology
of distributional trees and distributional forests following \cite{Schlosser+Hothorn+Stauffer:2019}. 
In that way no model assumptions are required as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while combining an 
ensemble of trees to a forest model allows for modeling smooth changes as well. 
Thus, a high level of flexibility is provided while a full distributional model is specified 
by distributional trees and forests such that a wide range of inference is made available.

\fixme{include and revise ``motivating'' example}

As a first illustration Figure~\ref{fig:tree_ibk} displays a distributional
tree applied to predict the wind direction at the airport of Innsbruck in the following hour. 
Innsbruck lies at the bottom of a deep valley in the Alps. Topography channels wind along the 
west-east valley axis or along a tributary valley intersecting from the south. Hence, pressure 
gradients to which valley wind regimes react both west and east of the airport are considered 
as covariates along with other meteorological measurements at the airport (lagged by one hour),
such as wind direction and wind speed at Innsbruck Airport.

\begin{figure}[t]
\centering
\includegraphics[width = \textwidth]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
\caption{Fitted tree based on the von Mises distribution for wind direction forecasting.
In each terminal node the empirical histogram (gray) and fitted density (red line)
are depicted along with the estimated location parameter (red hand). The covariates
employed are wind direction (degree), wind speed ($\text{ms}^{-1}$),
and pressure gradients ($\text{dpressure; hPa}$) west and east of the airport,
all lagged by one hour.}
\label{fig:tree_ibk} 
\end{figure}

The corresponding methodology and its features are presented in Section~\ref{sec:tree_forest} after
discussing distributional models for circular data in Section~\ref{sec:prob_circ}.
In Section~\ref{sec:wind} the presented models are applied to probabilistic forecasting of wind direction
at the airports of Innsbruck and Vienna.

\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper circular
probability distribution, i.e., a probability distribution which accounts for the periodicity
of circular data.
Generally, this feature can be obtained by ``wrapping'' the probability density function
of any continuous distribtuion around the unit circle \citep{Mardia+Jupp:2009}, for example by
applying the modulo operation on the interval length $2\pi$.
In that way, the wrapped Cauchy distribution or the wrapped normal distribution can be employed to model symmetric unimodal circular data.
A close approximation to the wrapped normal distribution that is mathematically simpler and hence easier 
to use \citep{Fisher:1993} is provided by the von Mises distribution,
a purely circular distribution which is also known as ``the circular normal distribution''
and commonly applied for probabilistic modeling of circular data.
Based on a location parameter $\mu \in (-\pi, \pi]$ and a concentration parameter $\kappa > 0$ 
\fixme{check domain of $\mu$ in literature}
the density of the von Mises distribution for an observation $y \in (-\pi, \pi]$ is given by:
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$
\citep[see, e.g.,][for a more detailed overview]{Jammalamadaka+Sengupta:2001}.

@@

\begin{figure}[t]
\centering
\includegraphics[width=4cm]{density_vM.pdf}
\caption{Illustration of a distributional model fitted to circular data by maximum likelihood-estimation
employing the von Mises distribution. Empirical histogram (gray) and fitted density (red line) are depicted
along with the estimated location parameter (red hand).
}
\label{fig:density_vM} 
\end{figure}
\fixme{update figure 2d to 3d, using data of example tree}

@@

\section{Circular regression trees and forests}
\label{sec:tree_forest}
\fixme{revise transition from sec\,2. to sect\,3. with the focus on circular methods}
A common way of including covariates as regressors in distributional models is to apply
generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} where each distribution parameter is estimated
by its own generalized additive model (GAM).
As already discussed in Section~\ref{sec:introduction}, additive strategies work well 
for smooth effects but difficulties can arise in case of non-additive changes. Additionally, 
specifying a model properly can be very challenging, particularly for a high number of covariates 
and no information on possible interactions.
An alternative regression approach that can deal with these problems is to apply a tree-structured 
model.

\subsection{Circular trees}
\label{sec:circtree}
Fitting a global model to a full data set can be very challenging, particularly for 
complex data with strong variations. Therefore, separating the data set into more homogeneous 
subgroups based on covariates before fitting a local model in each of these subgroups allows 
to capture (possibly) group specific effects more precisely and hence can result in a better 
fitting overall model.
This is the general idea of regression trees which are combined with distributional regression
modeling in \cite{Schlosser+Hothorn+Stauffer:2019} based on the unbiased recursive partitioning 
algorithms MOB \citep{Zeileis+Hothorn+Hornik:2008} or CTree \citep{Hothorn+Hornik+Zeileis:2006}.
By applying the presented distributional trees a full distributional model is specified in each 
node of the tree. For this purpose a distribution family has to be chosen in advance. Hence, 
selecting the von Mises distribution allows for an application of distributional trees to circular
data following the tree building algorithm as introduced in Section~2.2. of 
\cite{Schlosser+Hothorn+Stauffer:2019}.

@@

\subsection{Circular forests}
\label{sec:circforest}
A natural extension of distributional trees for circular data are ensembles or forests 
that can improve forecasts by regularizing and stabilizing the model.
Random forests introduced by \cite{Breiman:2001} average the predictions of an ensemble
of trees, each built on a subsample or bootstrap of the original data. 
A generalization of this strategy is to obtain weighted predictions by adaptive local 
likelihood estimation of the distributional parameters 
\citep[Section~2.3. of][]{Schlosser+Hothorn+Stauffer:2019, Hothorn+Zeileis:2017}. 
More specifically, for each possibly new observation~$\bm{z}$ a set of averaged ``nearest neighbor''
weights $w_i^{forest}(\bm{z})$ 
%\citep{Lin+Jeon:2006}
is obtained that is based on the number of trees in which $\bm{z}$ is assigned to the same terminal 
node as each learning observation $y_i, i \in \{1,\ldots,n\}$.
Hence, for a forest of $T$ trees the weights are calculated via
\begin{equation}
w^{\text{forest}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
\frac{\mathbf{1}((\bm{z}_i \in \mathcal{B}^t_b) \land (\bm{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
\end{equation}
where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
segment of the $t$-th tree.
Hence, similar observations ending up more often in the same terminal node have higher weights
and in that way a stronger influence in the estimation process.

@@

\section{Case study: Probabilistic wind direction forecasting}
\label{sec:wind}

@@

@@

\fixme{information on NAs, update data information, include overview}

@@

\subsection{Evaluation and models}
Along with distributional trees and forests for circular data three other models have been applied
in this study. In particular, they have been evaluated in a five-fold cross validation with each of 
the five years contained in the data set being used as testing data once after the models where 
learned on the remaining four years.

@@

\fixme{condense evaluation}

@@

\fixme{comment on aggregation}

@@

\fixme{condense model descriptions}

@@

@@

\fixme{add information on applied implementation here (!) or in separate chapter?}

\subsection{Results}
\fixme{update results section}
In order to test and compare the models over varying settings not only 1-hour but also 3-hour
forecasts are considered at both airports yielding four different scenarios of which the results
are discussed in this section.

@@

\begin{figure}[t]
\centering
\setkeys{Gin}{width=1.0\textwidth}
%\hspace{-2cm}
%\includegraphics{boxplot_crpsraw.pdf}
\includegraphics{_plot_circforest_validation_crpsraw_agg_comparison_with_lowff_v14b.pdf}
\caption{}
\label{fig:boxplot_crpsraw} 
\end{figure}

@@

\begin{figure}[t]
\centering
\setkeys{Gin}{width=1.0\textwidth}
%\hspace{-2cm}
%\includegraphics{boxplot_crpsskill.pdf}
\includegraphics{_plot_circforest_validation_crpsskill_agg_comparison_with_lowff_v14b.pdf}
\caption{}
\label{fig:boxplot_crpsskill} 
\end{figure}

\section{Summary and conclusion}
Distributional trees for circular responses are established by coupling
model-based recursive partitioning with the von Mises distribution.
The resulting trees can capture nonlinear changes, shifts, and potential interactions
in covariates without prespecification of such effects. This is particularly
useful for modeling wind direction in mountainous terrain where wind shifts
can occur due to turns of the pressure gradients along a valley.

\fixme{add summary of results}

\fixme{outlook on possible extensions of models, use of NWP data}
